Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-04-15 manu_draft.Rmd
more gams description

To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program, including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.  These methods leverage functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods. In generals, GAMs are an extension of linear models that can incorporate additive smoothed tersm to model non-linear effects. The basic structural form is (Hastie and Tibshirani 1986):

$$
E\left(Y|X_1, X_2, ..., X_p\right) = s_0 + \sum_{j=1}^{p} s_j\left(X_j\right) 
$$

where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an interecept term $s_0$.  The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.  Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.

The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.  A spline is a piecewise function, such as a polynomial, that is connected at knots that smoothly join the separate pieces of the function.  Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.  There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to esitmating the smoother.  The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).  In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty. This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function. However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.  The default is ten, and as described below, this default value was increased to prevent under-fitting the data.

An additional smoother provided by *mgcv* is the cubic spline, which provides similar functionality as the thin-plate spline, but has an explicit form as a cubic polynomial function that is continuous to the second derivative at each knot location. Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is a useful form for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.  A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of component being modelled.  This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.  In others words, not using a cyclic cubic spline to modeal season would cause a non-continous overall function with step increases or decreases for the seasonal estimates between the years.  This discrepancy can likely lead to inaccurate trend estimates when evaluating seasonal differences over time.

Because GAM structures are flexible, the functional form for time can be described wih varying levels of complexity depending on the desired structural form.  The method is useful for time series analysis because separate components can be explicitly modelled with differnet smoothing tersm.

using some of the model structures they have documented and incorporated into their R package, baytrends. Table 1 summarizes the four primary model structures tested. The first three models (gam0, gam1, gam2) have the same structure as CBP’s base models. We added a fourth model, gam6. For all models, date is expressed as cyear, or centered decimal date, meaning that a date is turned into a decimal (i.e., 2002.41), and then a time series is centered so that the middle date in a record becomes zero. doy is the day of year as a numeric value from 1 to 366. All models include year as a linear effect. The functions s() model either year or doy as a smoothed, non-linear variable and ti() models the interaction between the two in the gam2 model. A central challenge to optimizing GAMs for applications such as this, where time series data is rather noisy with significant inner- and interannual seasonal variation, is estimating how smooth (vs. ‘wiggly’) a spline should be to fit the data without over-fitting (i.e. following noise). In the mgcv package this is done by GCV. The fourth model, gam6, has the same structure as gam1 but permits a much higher degree in the s(year) by increasing the maximum allowable number of knots, k, and relies on mgcv’s internal model selection statistics to determine the appropriate degree of smoothness (using GCV).

>>>>

Generalized Additive Models are an extension of linear models that can incorporate additive smoothed terms to model non-linear effects. The basic structural form is (Hastie and Tibshirani 1986):

\begin{equation}
E\left(Y|X_1, X_2, ..., X_p\right) = s_0 + \sum_{j=1}^{p} s_j\left(X_j\right) 
(\#eq:gengam)
\end{equation}

where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an interecept term $s_0$.  The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.  Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.  All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods (Wood, 2017).

The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.  A spline is a piecewise function, such as a polynomial, that is connected at knots (or *k*-values) that smoothly join the separate pieces of the function.  Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.  There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother (Wood, 2003).  The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).  In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty. This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function. However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.  The default is ten, and as described below, this default value was increased to prevent under-fitting the data.

An additional smoother provided by *mgcv* is the cubic spline, which provides similar functionality as the thin-plate spline, but has an explicit form as a cubic polynomial function that is continuous to the second derivative at each knot location. Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is a useful form for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.  A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of component being modelled.  This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.  In others words, not using a cyclic cubic spline to model season would cause a non-continous function across the time series with step increases or decreases for the seasonal estimates between the years.  This discrepancy can likely lead to inaccurate or ambiguous trend estimates when evaluating seasonal differences over time.

To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program (Murphy et al. 2019), including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.  These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.  This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.  Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.  We present four GAM structures, following a similar format as Murphy et al. (2019):

\begin{equation}
\boldsymbol{gam0:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + \epsilon 
(\#eq:gamzero)
\end{equation}

\begin{equation}
\boldsymbol{gam1:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
(\#eq:gamone)
\end{equation}

\begin{equation}
\boldsymbol{gam2:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + ti_1(doy, year) + \epsilon
(\#eq:gamtwo)
\end{equation}

\begin{equation}
\boldsymbol{gam6:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2^*\left(year\right) + \epsilon
(\#eq:gamsix)
\end{equation}

For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $year$ and day of year ($doy$).  The $year$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).  The $doy$ variable is an integer value for the Julian day in each year (from 1 to 366) that represents the seasonal component of the time series. The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models. The $s_1()$ and $s_2()$ functions are different smoothers for $year$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component.  The $ti_1()$ function is a tensor product interaction between $year$ and $doy$ that allows the seasonal component to change between years.  Within the $ti_1()$ function are the two separate smoothers for year as a thin-plate spline and $doy$ as a cyclic cubic spline. As with conventional linear models, the remaining variance in the response not captured by the functional form of the model is expressed as an error term $\epsilon$ that is normally-distributed having mean equal to zero and variance as $\sigma^2$.  The residual variance is used to back-transform model estimates for trend testing, described below.

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameter by better satisying the model assumptions.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).  These methods... 
A comparison of fit between the two methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.

A central challenge to optimizing GAMs for applications such as this, where time series data is rather noisy with significant inner- and interannual seasonal variation, is estimating how smooth (vs. ‘wiggly’) a spline should be to fit the data without over-fitting (i.e. following noise). The four models above, Table 1.... In the mgcv package this is done by GCV. The fourth model, gam6, has the same structure as gam1 but permits a much higher degree in the s(year) by increasing the maximum allowable number of knots, k, and relies on mgcv’s internal model selection statistics to determine the appropriate degree of smoothness (using GCV).
